---
title: "Global Passport Strengths"
subtitle: "STA313 Spring 2026 - Project 1"
author: "Standard Deviants"
format:
  revealjs:
    slide-number: true
    slide-number-format: "{current}"
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(jsonlite)
```

```{r}
#| label: load-data
#| include: false
country_lists <- read_csv("data/country_lists.csv")
rank_by_year <- read_csv("data/rank_by_year.csv")
unsd <- read_csv2("data/UNSD_Methodology.csv")
```

## Henley Passport Index {.smaller}

Data distributed via **TidyTuesday**, prepared by Brenden Smith and Jen Richmond from the **Henley Passport Index** API by Henley & Partners.

- **country_lists.csv** — 199 passports with destinations broken down by access type: visa-free, visa on arrival, e-visa, ETA, and visa required
- **rank_by_year.csv** — yearly passport rankings for 199 countries from 2006 to 2025, including region and total visa-free destination count

We chose this dataset because we love traveling and wanted to better understand the travel restrictions faced by citizens around the world.

**Questions explored:**

1. Which countries are in the top 5 for the greatest gains in passport power, and which fall into the bottom 5 for the largest losses in power?
2. How does the average visa accessibility score vary by region, and which regions rank highest/lowest on this score?




## Top Gains and Losses in Visa-Free Travel {.smaller}

```{r}
#| label: quesstion1-plot1-data
#| include: false
df_passport <- read_csv("data/rank_by_year.csv") |>
  mutate(country = as.factor(country))
passport_plot_data <- df_passport |>
  dplyr::filter(year %in% c(2019, 2025)) |>
  group_by(country) |>
  dplyr::filter(n() == 2) |>
  ungroup() |>
  select(country, year, visa_free_count) |>
  pivot_wider(
    names_from = year,
    values_from = visa_free_count,
    names_prefix = "yr_"
  ) |>
  mutate(
    change_visa_free_count = yr_2025 - yr_2019,
    neg_pos = if_else(change_visa_free_count >= 0, "pos", "neg"),
    label_position = if_else(neg_pos == "pos", -0.5, 0.5)
  )

top_5 <- passport_plot_data |>
  slice_max(change_visa_free_count, n = 5, with_ties = FALSE)
bottom_5 <- passport_plot_data |>
  slice_min(change_visa_free_count, n = 5, with_ties = FALSE)
final_plot_df <- bind_rows(top_5, bottom_5) |>
  distinct(country, .keep_all = TRUE) |>
  mutate(country = fct_reorder(country, change_visa_free_count))
```

```{r}
#| label: question1plot1
#| fig-width: 10
#| fig-asp: 0.6
#| fig-align: center
ggplot(final_plot_df, aes(y = country, x = change_visa_free_count)) +
  geom_vline(xintercept = 0) +
  geom_col(aes(fill = neg_pos), show.legend = FALSE) +
  geom_text(
    aes(
      x = 0,
      label = str_remove(country, " \\(.*\\)"),
      color = neg_pos,
      hjust = if_else(neg_pos == "pos", 1.1, -0.1)
    ),
    show.legend = FALSE,
    fontface = "bold"
  ) +
  geom_text(
    aes(
      label = paste0(
        if_else(change_visa_free_count > 0, "+", ""),
        change_visa_free_count
      ),
      hjust = if_else(neg_pos == "pos", 1.2, -0.2)
    ),
    color = "white",
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("neg" = "#800000", "pos" = "#001f3f")) +
  scale_color_manual(values = c("neg" = "#800000", "pos" = "#001f3f")) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  labs(
    y = NULL,
    title = "Passport Power Shifts (2019–2025)",
    subtitle = "Top 5 Gains and Losses",
    caption = "Source: Henley Passport Index rank_by_year.csv",
    x = "Changes in Visa Free Count"
  )

```

## Average Yearly Changes \n in Visa-Free Travel {.smaller}

```{r}
#| label: q1plot2-data
#| include: false

ranks_by_year <- df_passport |>
  filter((country %in% final_plot_df$country) & between(year, 2019, 2025)) |>
  group_by(country)

top_5_avg <- ranks_by_year |>
  slice_max(visa_free_count, n = 5, with_ties = FALSE) |>
  group_by(year) |>
  summarize(
    year_average_top = mean(visa_free_count)
  )
bottom_5_avg <- ranks_by_year |>
  slice_min(visa_free_count, n = 5, with_ties = FALSE) |>
  group_by(year) |>
  summarize(
    year_average_bottom = mean(visa_free_count)
  )

rank_by_year_final <- bind_rows(top_5_avg, bottom_5_avg) |>
  pivot_longer(
    cols = c(year_average_top, year_average_bottom),
    names_to = "section",
    values_to = "average"
  ) |>
  drop_na() |>
  mutate(
    section = fct_relevel(section, c("year_average_top", "year_average_bottom"))
  )


```

```{r}
#| label: q1plot2
#| fig-width: 10
#| fig-asp: 0.4
#| fig-align: center

ggplot(
  rank_by_year_final,
  aes(x = year, y = average, group = section, color = section, fill = section)
) +
  geom_line(linewidth = 0.8) +
  geom_point(color = "white", shape = 21, size = 3, show.legend = FALSE) +
  scale_color_manual(
    values = c(
      "year_average_bottom" = "#800000",
      "year_average_top" = "#001f3f"
    ),
    labels = c(
      "year_average_bottom" = "Bottom 5 Average",
      "year_average_top" = "Top 5 Average"
    )
  ) +
  scale_fill_manual(
    values = c(
      "year_average_bottom" = "#800000",
      "year_average_top" = "#001f3f"
    )
  ) +
  scale_x_continuous(
    breaks = seq(2019, 2025, by = 1)
  ) +
  labs(
    x = "Year",
    color = "Section",
    fill = "Section",
    title = "Average changes in the Visa Free Count for the Top 5 and Bottom 5 Countries",
    subtitle = "2019-2025",
    caption = "Source: Henley Passport Index rank_by_year.csv",
    y = "Visa Free Count"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

## Visa Accessibility by Sub-region Over Time {.smaller}

```{r}
#| label: plot1-data
#| include: false

# cleaning unsd
unsd_clean <- unsd |>
 select(iso2 = `ISO-alpha2 Code`, subregion = `Sub-region Name`) |>
 filter(!is.na(iso2), iso2 != "")

# Namibia, Kosovo, and Taiwan didnt have pairings to a subregion for some reason
extra_mappings <- tibble(
 iso2      = c("XK", "TW", "NA"),
 subregion = c("Eastern Europe", "Eastern Asia", "Sub-Saharan Africa")
)

unsd_clean <- bind_rows(unsd_clean, extra_mappings)

years_to_show <- c(2006, 2010, 2014, 2018, 2022, 2025)
legend_order <- c(
 "Western Europe",
 "Northern Europe",
 "Southern Europe",
 "Eastern Europe",
 "Western Asia",
 "Central Asia",
 "Southern Asia",
 "Eastern Asia",
 "South-eastern Asia",
 "Northern Africa",
 "Sub-Saharan Africa",
 "Northern America",
 "Latin America and the Caribbean",
 "Australia and New Zealand"
)
# Give subregions their scores, leave out Melanesia, Micronesia, Polynesia for vis reasons
subregion_rank <- rank_by_year |>
 filter(year %in% years_to_show) |>
 left_join(unsd_clean, by = c("code" = "iso2")) |>
 filter(!is.na(subregion), !subregion %in% c("Melanesia", "Micronesia", "Polynesia")) |>
 group_by(subregion, year) |>
 summarise(avg_score = mean((1 - rank / 116) * 100, na.rm = TRUE), .groups = "drop") |>
 mutate(subregion = factor(subregion, levels = legend_order))

```

```{r}
#| label: plot1
#| fig-width: 10
#| fig-asp: 0.6
#| fig-align: center
subregion_rank |>
 ggplot(aes(x = year, y = avg_score, colour = subregion, group = subregion)) +
 geom_line(size = 1) +
 geom_point(size = 2) +
 scale_colour_viridis_d(breaks = legend_order, option = "turbo", begin = 0, end = 1) +
 scale_x_continuous(
   breaks = years_to_show
 ) +
 scale_y_continuous(
 limits = c(0, 100),
 breaks = seq(0, 100, by = 20)
 ) +
 labs(
   x = "Year",
   y = "Average Accessibility Score (0-100)",
   colour = "Subregion",
   title = "Visa Accessibility by Sub-region Over Time",
   subtitle = "Average accessibility score per subregion (2006–2025), higher score = greater visa-free access",
   caption = "Sources: Passport Index rank_by_year.csv, UN Statistics Division (UNSD) subregion classification"
 ) +
 theme_minimal() +
 theme(
   plot.title = element_text(face = "bold", size = 15),
   plot.caption = element_text(size = 8, color = "black", hjust = 1),
   plot.caption.position = "plot",
   legend.title = element_text(face = "bold")
 )
```

## Visa Access Type Composition by Region (2025) {.smaller}

```{r}
#| label: plot2-data
#| include: false
#| # JSON parsing
region_map <- rank_by_year |>
 filter(year == 2025) |>
 select(code, region)

# Helper: safely parse JSON array string -> count of destinations
count_destinations <- function(x) {
 parsed <- jsonlite::fromJSON(x)
 sum(sapply(parsed, nrow))
}

region_access <- country_lists |>
 select(code, visa_required, visa_online, visa_on_arrival,
        visa_free_access, electronic_travel_authorisation) |>
 pivot_longer(
   cols = -code,
   names_to = "access_type",
   values_to = "json_val"
 ) |>
 mutate(
   access_type = recode(access_type,
     "electronic_travel_authorisation" = "eta",
     "visa_free_access" = "visa_free"
   ),
   access_type = factor(access_type, levels = c(
     "visa_required", "visa_online", "visa_on_arrival", "eta", "visa_free"
   )),
   n_destinations = map_int(json_val, count_destinations)
 ) |>
 # Join to get region
 inner_join(region_map, by = "code") |>
 # Aggregate to region level
 group_by(region, access_type) |>
 summarise(avg_destinations = mean(n_destinations, na.rm = TRUE), .groups = "drop")

region_order <- region_access |>
 filter(access_type == "visa_free") |>
 arrange(avg_destinations) |>
 pull(region)

region_access <- region_access |>
 mutate(region = factor(region, levels = region_order))

```

```{r}
#| label: plot2
#| fig-width: 10
#| fig-asp: 0.6
#| fig-align: center
region_access |>
 ggplot(aes(x = region, y = avg_destinations, fill = access_type)) +
 geom_col(position = "fill") +
 scale_fill_viridis_d(
   option = "D",
   labels = c(
     eta = "ETA",
     visa_free = "Visa Free",
     visa_on_arrival = "Visa on Arrival",
     visa_online = "Visa Online",
     visa_required = "Visa Required"
   )
 ) +
 scale_y_continuous(labels = scales::percent) +
 labs(
   x = "Region",
   y = "Share of Destinations (%)",
   fill = "Access Type",
   title = "Visa Access Type Composition by Region (2025)",
   subtitle = "Proportion of destinations per access type, averaged across passports in each region",
   caption = "Sources: Passport Index country_lists.csv, rank_by_year.csv"
 ) +
 theme_minimal() +
 theme(
   plot.title = element_text(face = "bold", size = 15),
   plot.caption = element_text(size = 8, color = "black", hjust = 1),
   plot.caption.position = "plot",
   legend.title = element_text(face = "bold")
 )
```
